Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-25 paper/zhang-cook-laa-langrene-menendez.Rmd
indices -> index, australian english, some gramma issues

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu 
abstract: A guided tour helps to visualise high-dimensional data by showing the low-dimensional projections along a projection pursuit optimization path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data.  While much work has been done in developing new indexes in the literature, less has been done on understanding the optimization. Index functions can be noisy, they might have multiple local maxima as well as an optimal maximum, and it is constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package, \pkg{ferrn}.
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

The focus of this paper is on the optimisation problem arising in the projection pursuit guided tour [@buja2005computational], an exploratory data analysis technique used for detecting interesting structures in high-dimensional data through a set of lower-dimensional projections [@cook2008grand]. The goal of the optimisation is to identify the projection, represented by the projection matrix, that gives the most interesting low-dimensional view. A view is interesting if it can show some structures of the data that depart from normalality, such as bimodality, clustering, or outliers.

@@

Random search methods [@Romeijn2009; @zabinsky2013stochastic; @andradottir2015review] have a random sampling component as part of their algorithm and have been shown to have the ability to optimise non-convex and non-smooth functions. The initial random search algorithm, pure random search [@Brooks1958-zk], draws candidate points from the entire space without using any information of the current position and updates the current position when an improvement on the objective function is made. As the dimension of the space becomes larger, sufficient sampling from the entire space would require a long time for convergence to occur, despite a guaranteed global convergence [@spall2005introduction]. Various algorithms have thus been developed to improve pure random search by either concentrating on a narrower sampling space or using a different updating machanism. Creeping random search [@White1971-ci] is such a variation, where a candidate point is generated within a neighbourhood of the current point. This makes creeping random search faster to compute, however, global convergence is not longer guaranteed. On the other hand, simulated annealing [@kirkpatrick1983optimization; @bertsimas1993simulated], introduces a different updating mechanism. Rather than only updating the current point when an improvement is made, simulated annealing uses a Metropolis acceptance criterion, where worse candidates still have a chance to be accepted. The convergence of simulated annealing algorithms has been widely researched [@mitra1986convergence; @granville1994simulated] and the global optimum can be attained under mild regularity conditions.

Pseudo derivative search uses a common search scheme in optimistaion: line search. In line search methods, users are required to provide an initial estimate $x_{1}$ and, at each iteration, a search direction $S_k$ and a step size $\alpha_k$ are generated. Then one moves on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and the process is repeated until the desired convergence is reached. In derivative-free methods, local information of the objective function is used to determine the search direction. The choice of step size also needs consideration, as inadequate step sizes might prevent the optimisation method to converge to an optimum. An ideal step size can be chosen by finding the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ at each iteration.

@@

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. An interesting projection shows structures that are non-normal since theoretical proofs from @diaconis1984asymptotics have shown that projections tend to be normal, as $n$ and $p$ approach infinity under certain conditions. There has been plentiful index functions proposed in the literature and here list a few: early indices that can be categorised as measuring the $L^2$ distance between the projection and a normal distribution: Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], and natural Hermite index [@cook1993projection]; chi-square index [@posse1995projection] for detecting spiral structure; LDA index[@lee2005projection] and PDA [@lee2010projection] index for supervised classification; kurtosis index [@Loperfido2020] and skewness index [@Loperfido2018] for detecting outliers in financial time series; and most recently, scagnostic indices [@laa2020using] for summarising structures in scatterplot matrices based on eight scagnostic measures.

@@

Simulated annealing (SA) uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current. Given an initial value of $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SA gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even if the candidate basis does not have a higher index value and hence enable the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} in the appendix highlights how SA differs from CRS in the inner loop.

@@

A starting point of diagnosing an optimiser is to understand how many searches it has conducted, i.e. we want to summarise how the index is increasing over iterations and how many basis need to be sampled at each iteration. This is achieved using the function `explore_trace_search()`: a boxplot shows the distribution of index values for each try, where the accepted basis (corresponding to the highest index value) is always shown as a point. In the case of few tries at a given iteration, showing the data points directly may be preferred over the boxplot, this is controlled via the `cutoff` argument. Additional annotations are added to facilitate better reading of the plot and these include 1) the number of points searched in each iteration can be added as text label at the bottom of each iteration; 2) the anchor bases to interpolate are connected and highlighted in a larger size; and 3) the colour of the last iteration is in a grey scale to indicate no better basis found in this iteration.

@@

```{r toy-interp, fig.cap = "An inspection of the index values as the optimisation progress for two optimisers: CRS (left) and SA (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \\code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SA is observed to first pass by some bases with higher index value before reach the target bases in time 76-130."}
```

Another interesting feature to examine is the changes on the index value between interpolating bases since the projection on these bases is shown in the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisations as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at the first sight that the interpolation sometimes passes bases with higher index value before it decreases to a lower target. This happens because on the one hand, the probabilistic acceptance in SA implies that some worse bases will be accepted by the optimiser. In addition, the guided tour interpolates between the current and target basis to provide a smooth transition between projections, and sometimes a higher index value will be observed along the interpolation path. This indicates that a non-monotonic interpolation can't be avoided, even for CRS. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for CRS.

@@

```{r toy-pca, fig.height = 5, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD has been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is also of interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a Stiefel manifold. For one-dimensional projections this forms a $p$-dimensional sphere. A dimensionality reduction method, e.g. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various types of bases involved: 1) The starting basis; 2) The search bases that the optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally, 5) the end basis. The importance of these bases differs and the most important ones are the starting, interpolating and end bases. The anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the orientation of the basis and so is the index value. However, this creates difficulties for comparing the optimisers since the end bases will be symmetric with respect to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high-dimensional sphere and random bases on the sphere can be generated via \CRANpkg{geozoo} [@geozoo] package. Along with the bases recorded during the optimisation and a zero basis, the parameters (centre and radius) of the 2D space can be estimated. PCA is performed to get the first two PC coordinates of all the bases and the centre of the circle is the PC coordinates of the zero matrix, and radius is estimated as the largest distance from the centre to the basis. The theoretical best basis is known for simulated data and can be labelled to compare how close the end basis found by the optimisers is to the theoretical best. Various aesthetics, i.e. size, alpha and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha is used. Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

Figure \ref{fig:toy-pca} shows the PCA plot of CRS and PD for a 1D projection problem. Both optimisers find the optimum but PD gets closer. With the PCA plot, one can visually appreciate the nature of these two optimisers: PD first evaluates points in a small neighbourhood for a promising direction, while CRS evaluates points randomly in the search space to search for the next target. There are dashed lines annotated for CRS and it describes the interruption of the interpolation, which will be discussed in details in Section \ref{monotonic}.

@@

```{r toy-pca-animated, fig.cap = "Six frames selected from the animated version of the previous plot. With animation, the progression of the search paths from start to finish are better identified. CRS (green) finishes the optimisation quicker than PD (brown) since there is no further movement for CRS in the sixth frame. The full video of the animation can be found at \\url{https://vimeo.com/504242845}."}
```

@@

Both optimisers have found the third variable in the first direction, and the second variable in the second direction. Note however the different orientation of the basis, following from the different sign in the second column. One would expect to see this in the torus plot as the final bases match each other when projected onto one torus circle (due to the same sign in the first column) and be symmetric when projected onto the other (due the the sign difference in the second column). In Figure \ref{fig:toy-torus}, this can be seen most clearly in frame 5 where the two circles are rotated into a line from our view.

```{r toy-torus, fig.cap="Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and CRS (green). The projection problem is a 2D projection with three variables using holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \\url{https://vimeo.com/512882784}."}
```

@@

Random variables with different distributions have been simulated and the distributional form of each variable is presented in equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variables `x1`, `x8`, `x9` and `x10` are normal noise with zero mean and unit variance and `x2` to `x7` are normal mixtures with varied weights and locations. All the variables have been scaled to have an overall unit variance before projection pursuit. Holes index [@cook2008grand], used for detecting binormality of the variables, is used throughout the examples unless otherwise specified.

@@

```{r polish, fig.height= 5, fig.cap = "Comparison of the projected data before and after using polishing for a 2D projection problem on \\code{boa6} data using holes index. Top row shows the intial projected data and the final views after CRS and polish search and the second row traces the index value. The clustering structure in the data is detected by CRS (top middle panel) but the polish step improves the index value and produces clearer boundaries of the clusters (top right panel), especially along the vertial axis."}
```

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to see if an even better basis can be found. A polish search takes the final basis of an optimiser as the start of a new guided tour to search for local improvements. The polish algorithm is similar to the CRS but with three distinctions: 1) a hundred rather than one candidate bases are generated each time in the inner loop; 2) the neighbourhood size is reduced in the inner loop, rather than in the outer loop; and 3) three more termination conditions have been added to ensure the new basis generated is distinguishable from the current one in terms of the distance in the space,  percentage change in the index value, and neighbourhood size:

1) the distance between the basis found and the current basis needs to be larger than 1e-3;
2) the percentage change of the index value need to be larger than 1e-5; and
3) the alpha parameter on itself needs to be larger than 0.01

Figure \ref{fig:polish} presents the projectd data and trace plot of a 2D projection, optimised by CRS and followed by the polish step. The top row shows the intial projection, the final projection after CRS, and the final projection after polish, repsectively. The end basis found by CRS reveals the four clusters in the data, but the edges of each cluster are not clean-cut. Polish works with this end basis and further pushes the index value to produce much clear edges of the cluster, especially along the vertical axis.

@@

The holes index function used for all the examples before this section produces a smooth interpolation, while this is not the case for all the indices. An example of a noisy index function for 1D projections compares the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, using the Kolmogorov test. Let $F_{.}(n)$ be the ECDF function, with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov index, $I^{nk}(n)$, is defined as:

@@

```{r noisy-better-geo, fig.height = 5, fig.cap = "Comparison of the three optimisers in optimising $I^{nk}(n)$ index for a 1D projection problem on a five-variable dataset, \\code{boa5}. Both CRS and SA succeeds in the optimisation, PD fails to optimise this non-smooth index. Further, SA takes  much longer than CRS to finish the optimisation, it finishes off closer to the theoretical best."}
```

@@

```{r kol-result, fig.height = 4, fig.cap="Comparing 20 search paths in the PCA-projected basis space facetted by two optimisers: CRS and SA, and two search sizes: 0.5 and 0.7. The optimisation is on the 1D projecton index, $I^{nk}(n)$, for \\code{boa6} data, where a local optimum, annotated by the cross (x), is presented in this experiment, along with the global optimum (*)."}
```

The next experiment compares the performance of CRS and SA when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With CRS and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing CRS and SA for a search size of 0.5, SA does not seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SA is working hard to examine if there is any other optimum in the basis space but the relative small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SA, with a reasonable large search window, is able to overcome the local optimum and optimise close towards the global optimum.

@@

One interesting situation observed in the previous examples is that, for some simulations as shown on the left panel of Figure \ref{fig:flip-sign}, the target basis is generated on the other half of the basis space and the interpolator seems to draw a straight line to interpolate. As mentioned previously, bases with opposite signs do not affect the projection and index value, but clearly, we prefer the target to have the same orientation as the current basis. The orientation of two bases can be checked via calculating the determinant and a negative value indicates opposite direction. Hence an orientation check is carried out once a new target basis is generated and the sign in the target basis will be flipped if a negative determinant is obtained. The interpolation after implementing the orientation check is shown on the right panel of Figure \ref{fig:flip-sign} where the unsatisfactory interpolation no longer exist.

@@

- Utilities
  - \code{theme\_fern()} and \code{format\_label()} for better display of the grid lines and axis formatting
  - \code{clean\_method()} to clean up the name of the optimisers
  - \code{botanical\_palettes} is a collection of colour palettes from Australian native plants. Quantitative palettes include daisy, banksia and cherry and sequential palettes contain fern and acacia
  - \code{botanical\_pal()} as the colour interpolator 
  - \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the colour and fill of the plot

>>>>

---
title: "Visual Diagnostics for Constrained Optimisation with Application to Guided Tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: CSIRO Data61 
    address: 34 Village Street, Docklands VIC 3008 Australia
    email: nicolas.langrene@csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu 
abstract: A guided tour helps to visualise high-dimensional data by showing the low-dimensional projections along a projection pursuit optimisation path. Projection pursuit is a generalisation of principal component analysis, in the sense that different indexes are used to define the interestingness of the projected data. While much work has been done in developing new indexes in the literature, less has been done on understanding the optimisation. Index functions can be noisy, they might have multiple local maxima as well as an optimal maximum, and it is constrained to generate orthonormal projection frames, which complicates the optimization. In addition, projection pursuit is primarily used for exploratory data analysis, and finding the local maxima is also useful. The guided tour is especially useful for exploration, because it conducts geodesic interpolation connecting steps in the optimisation and shows how the projected data changes as a maxima is approached. This work provides new visual diagnostics for examining a choice of optimisation procedure, based on the provision of a new data object which collects information throughout the optimisation. It has helped to diagnose and fix several problems with projection pursuit guided tour. This work might be useful more broadly for diagnosing optimisers, and comparing their performance. The diagnostics are implemented in the R package, \pkg{ferrn}.
keywords: graphics, multivariate, optimization
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

The focus of this paper is on the optimisation problem arising in the projection pursuit guided tour [@buja2005computational], an exploratory data analysis technique used for detecting interesting structures in high-dimensional data through a set of lower-dimensional projections [@cook2008grand]. The goal of the optimisation is to identify the projection, represented by the projection matrix, that gives the most interesting low-dimensional view. A view is interesting if it can show some structures of the data that depart from normality, such as bimodality, clustering, or outliers.

@@

Random search methods [@Romeijn2009; @zabinsky2013stochastic; @andradottir2015review] have a random sampling component as part of their algorithm and have been shown to have the ability to optimise non-convex and non-smooth functions. The initial random search algorithm, pure random search [@Brooks1958-zk], draws candidate points from the entire space without using any information of the current position and updates the current position when an improvement on the objective function is made. As the dimension of the space becomes larger, sufficient sampling from the entire space would require a long time for convergence to occur, despite a guaranteed global convergence [@spall2005introduction]. Various algorithms have thus been developed to improve pure random search by either concentrating on a narrower sampling space or using a different updating mechanism. Creeping random search [@White1971-ci] is such a variation, where a candidate point is generated within a neighbourhood of the current point. This makes creeping random search faster to compute, however, global convergence is not longer guaranteed. On the other hand, simulated annealing [@kirkpatrick1983optimization; @bertsimas1993simulated], introduces a different updating mechanism. Rather than only updating the current point when an improvement is made, simulated annealing uses a Metropolis acceptance criterion, where worse candidates still have a chance to be accepted. The convergence of simulated annealing algorithms has been widely researched [@mitra1986convergence; @granville1994simulated] and the global optimum can be attained under mild regularity conditions.

Pseudo derivative search uses a common search scheme in optimisation: line search. In line search methods, users are required to provide an initial estimate $x_{1}$ and, at each iteration, a search direction $S_k$ and a step size $\alpha_k$ are generated. Then one moves on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and the process is repeated until the desired convergence is reached. In derivative-free methods, local information of the objective function is used to determine the search direction. The choice of step size also needs consideration, as inadequate step sizes might prevent the optimisation method to converge to an optimum. An ideal step size can be chosen by finding the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ at each iteration.

@@

Let $\mathbf{X}_{n \times p}$ be the data matrix with $n$ observations in $p$ dimensions. A $d$-dimensional projection is a linear transformation from $\mathbb{R}^p$ into $\mathbb{R}^d$ defined as $\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{Y}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection matrix. We define $f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}$ to be an index function that maps the projected data $\mathbf{Y}$ onto a scalar value. This is commonly known as the projection pursuit index function, or just index function, and is used to measure the "interestingness" of a given projection. An interesting projection shows structures that are non-normal since theoretical proofs from @diaconis1984asymptotics have shown that projections tend to be normal, as $n$ and $p$ approach infinity under certain conditions. There has been plentiful index functions proposed in the literature and here list a few: early indexes that can be categorised as measuring the $L^2$ distance between the projection and a normal distribution: Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], and natural Hermite index [@cook1993projection]; chi-square index [@posse1995projection] for detecting spiral structure; LDA index[@lee2005projection] and PDA [@lee2010projection] index for supervised classification; kurtosis index [@Loperfido2020] and skewness index [@Loperfido2018] for detecting outliers in financial time series; and most recently, scagnostic indexes [@laa2020using] for summarising structures in scatterplot matrices based on eight scagnostic measures.

@@

Simulated annealing (SA) uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current. Given an initial value of $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SA gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even if the candidate basis does not have a higher index value and hence enables the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} in the appendix highlights how SA differs from CRS in the inner loop.

@@

A starting point of diagnosing an optimiser is to understand how many searches it has conducted, i.e. we want to summarise how the index is increasing over iterations and how many basis need to be sampled at each iteration. This is achieved using the function `explore_trace_search()`: a boxplot shows the distribution of index values for each try, where the accepted basis (corresponding to the highest index value) is always shown as a point. In the case of few tries at a given iteration, showing the data points directly may be preferred over the boxplot, this is controlled via the `cutoff` argument. Additional annotations are added to facilitate better reading of the plot and these include 1) the number of points searched in each iteration can be added as text label at the bottom of each iteration; 2) the anchor bases to interpolate are connected and highlighted in a larger size; and 3) the colour of the last iteration is in a greyscale to indicate no better basis found in this iteration.

@@

```{r toy-interp, fig.cap = "An inspection of the index values as the optimisation progress for two optimisers: CRS (left) and SA (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \\code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SA is observed to first pass by some bases with higher index value before reaching the target bases in time 76-130."}
```

Another interesting feature to examine is the changes in the index value between interpolating bases since the projection on these bases is shown in the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisations as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at the first sight that the interpolation sometimes passes bases with higher index value before it decreases to a lower target. This happens because, on the one hand, the probabilistic acceptance in SA implies that some worse bases will be accepted by the optimiser. In addition, the guided tour interpolates between the current and target basis to provide a smooth transition between projections, and sometimes a higher index value will be observed along the interpolation path. This indicates that a non-monotonic interpolation can't be avoided, even for CRS. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for CRS.

@@

```{r toy-pca, fig.height = 5, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD have been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

Apart from checking the search and progression of an optimiser, looking at where the bases are positioned in the basis space is also of interest. Given the orthonormality constraint, the space of projection bases $\mathbf{A}_{p \times d}$ is a Stiefel manifold. For one-dimensional projections, this forms a $p$-dimensional sphere. A dimensionality reduction method, e.g. principal component analysis is applied to first project all the bases onto a 2D space. In a projection pursuit guided tour optimisation, there are various types of bases involved: 1) The starting basis; 2) The search bases that the optimiser evaluated to produce the anchor bases; 3) The anchor bases that have the highest index value in each iteration; 4) The interpolating bases on the interpolation path; and finally, 5) the end basis. The importance of these bases differs and the most important ones are the starting, interpolating, and end bases. The anchor and search bases can be turned on with argument `details = TRUE`. Sometimes, two optimisers can start with the same basis but finish with bases of opposite sign. This happens because the projection is invariant to the orientation of the basis and so is the index value. However, this creates difficulties for comparing the optimisers since the end bases will be symmetric to the origin. A sign flipping step is conducted to flip the signs of all the bases in one routine if different optimisations finish at opposite places.

Several annotations have been made to help understanding this plot. As mentioned previously, the original basis space is a high-dimensional sphere and random bases on the sphere can be generated via \CRANpkg{geozoo} [@geozoo] package. Along with the bases recorded during the optimisation and a zero basis, the parameters (centre and radius) of the 2D space can be estimated. PCA is performed to get the first two PC coordinates of all the bases and the centre of the circle is the PC coordinates of the zero matrix, and radius is estimated as the largest distance from the centre to the basis. The theoretical best basis is known for simulated data and can be labelled to compare how close the end basis found by the optimisers is to the theoretical best. Various aesthetics, i.e. size, alpha, and colour, are applicable to emphasize critical elements and adjust for the presentation. For example, anchor points and search points are less important and hence a smaller size and alpha are used. Alpha can also be applied on the interpolation paths to show the start to finish from transparent to opaque.

Figure \ref{fig:toy-pca} shows the PCA plot of CRS and PD for a 1D projection problem. Both optimisers find the optimum but PD gets closer. With the PCA plot, one can visually appreciate the nature of these two optimisers: PD first evaluates points in a small neighbourhood for a promising direction, while CRS evaluates points randomly in the search space to search for the next target. There are dashed lines annotated for CRS and it describes the interruption of the interpolation, which will be discussed in detail in Section \ref{monotonic}.

@@

```{r toy-pca-animated, fig.cap = "Six frames selected from the animated version of the previous plot. With animation, the progression of the search paths from start to finish is better identified. CRS (green) finishes the optimisation quicker than PD (brown) since there is no further movement for CRS in the sixth frame. The full video of the animation can be found at \\url{https://vimeo.com/504242845}."}
```

@@

Both optimisers have found the third variable in the first direction, and the second variable in the second direction. Note however the different orientation of the basis, following from the different sign in the second column. One would expect to see this in the torus plot as the final bases match each other when projected onto one torus circle (due to the same sign in the first column) and are symmetric when projected onto the other (due the sign difference in the second column). In Figure \ref{fig:toy-torus}, this can be seen most clearly in frame 5 where the two circles are rotated into a line from our view.

```{r toy-torus, fig.cap="Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and CRS (green). The projection problem is a 2D projection with three variables using the holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \\url{https://vimeo.com/512882784}."}
```

@@

Random variables with different distributions have been simulated and the distributional form of each variable is presented in equations \ref{eq:sim-norm} to \ref{eq:sim-x7}. Variables `x1`, `x8` to `x10` are normal noise with zero mean and unit variance and `x2` to `x7` are normal mixtures with varied weights and locations. All the variables have been scaled to have an overall unit variance before projection pursuit. Holes index [@cook2008grand], used for detecting bimodality of the variables, is used throughout the examples unless otherwise specified.

@@

```{r polish, fig.height= 5, fig.cap = "Comparison of the projected data before and after using polishing for a 2D projection problem on \\code{boa6} data using holes index. The top row shows the initial projected data and the final views after CRS and polish search and the second row traces the index value. The clustering structure in the data is detected by CRS (top middle panel) but the polish step improves the index value and produces clearer boundaries of the clusters (top right panel), especially along the vertical axis."}
```

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to see if an even better basis can be found. A polish search takes the final basis of an optimiser as the start of a new guided tour to search for local improvements. The polish algorithm is similar to the CRS but with three distinctions: 1) a hundred rather than one candidate bases are generated each time in the inner loop; 2) the neighbourhood size is reduced in the inner loop, rather than in the outer loop; and 3) three more termination conditions have been added to ensure the new basis generated is distinguishable from the current one in terms of the distance in the space, the percentage change in the index value, and neighbourhood size:

1) the distance between the basis found and the current basis needs to be larger than 1e-3;
2) the percentage change of the index value needs to be larger than 1e-5; and
3) the alpha parameter on itself needs to be larger than 0.01

Figure \ref{fig:polish} presents the projected data and trace plot of a 2D projection, optimised by CRS and followed by the polish step. The top row shows the initial projection, the final projection after CRS, and the final projection after polish, respectively The end basis found by CRS reveals the four clusters in the data, but the edges of each cluster are not clean-cut. Polish works with this end basis and further pushes the index value to produce clearer edges of the cluster, especially along the vertical axis.

@@

The holes index function used for all the examples before this section produces a smooth interpolation, while this is not the case for all the indexes An example of a noisy index function for 1D projections compares the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, using the Kolmogorov test. Let $F_{.}(n)$ be the ECDF function, with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov index, $I^{nk}(n)$, is defined as:

@@

```{r noisy-better-geo, fig.height = 5, fig.cap = "Comparison of the three optimisers in optimising $I^{nk}(n)$ index for a 1D projection problem on a five-variable dataset, \\code{boa5}. Both CRS and SA succeed in the optimisation, PD fails to optimise this non-smooth index. Further, SA takes  much longer than CRS to finish the optimisation, it finishes off closer to the theoretical best."}
```

@@

```{r kol-result, fig.height = 4, fig.cap="Comparing 20 search paths in the PCA-projected basis space facetted by two optimisers: CRS and SA, and two search sizes: 0.5 and 0.7. The optimisation is on the 1D projection index, $I^{nk}(n)$, for \\code{boa6} data, where a local optimum, annotated by the cross (x), is presented in this experiment, along with the global optimum (*)."}
```

The next experiment compares the performance of CRS and SA when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by the optimiser and search size. With CRS and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing CRS and SA for a search size of 0.5, SA does not seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SA is working hard to examine if there is any other optimum in the basis space but the relatively small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SA, with a reasonable large search window, is able to overcome the local optimum and optimise close towards the global optimum.

@@

One interesting situation observed in the previous examples is that, for some simulations, as shown on the left panel of Figure \ref{fig:flip-sign}, the target basis is generated on the other half of the basis space and the interpolator seems to draw a straight line to interpolate. As mentioned previously, bases with opposite signs do not affect the projection and index value, but clearly, we prefer the target to have the same orientation as the current basis. The orientation of two bases can be checked via calculating the determinant and a negative value indicates opposite direction. Hence an orientation check is carried out once a new target basis is generated and the sign in the target basis will be flipped if a negative determinant is obtained. The interpolation after implementing the orientation check is shown on the right panel of Figure \ref{fig:flip-sign} where the unsatisfactory interpolation no longer exists.

@@

- Utilities
  - \code{theme\_fern()} and \code{format\_label()} for better display of the grid lines and axis formatting
  - \code{clean\_method()} to clean up the name of the optimisers
  - \code{botanical\_palettes} is a collection of colour palettes from Australian native plants. Quantitative palettes include daisy, banksia, and cherry and sequential palettes contain fern and acacia
  - \code{botanical\_pal()} as the colour interpolator 
  - \code{scale\_color\_*()} and \code{scale\_fill\_*()} for scaling the colour and fill of the plot
